Quality Criteria for SKOS Vocabularies

### 2012-03-21 criteria.tex
finalize labelling / documentation criteria

\begin{definition}[SKOS Vocabulary] Let a SKOS vocabulary be a tuple of the form $V = \langle IR, C, SR, ULV \rangle$, with \(IR = I_{CEXT}(\texttt{rdfs:Resource}^\mathcal{I})\) being the set of \textbf{resources}, \(C = I_{CEXT}(\texttt{skos:Concept}^\mathcal{I})\) being the set of \textbf{conceptual resources}, $ULV \subseteq I_{CEXT}(\texttt{rdfs:Literal}^\mathcal{I})$ being the set of \textbf{untyped literal values}, and \(SR = I_{EXT}(\texttt{skos:semanticRelation}^\mathcal{I})\) being the set of \textbf{semantic relations} associating conceptual resources with one another. Further, we let $V$ be the fully entailed RDFS interpretation~\cite{RDFSEM2012} of the underlying RDF graph, enriched by entailment of \texttt{owl:inverseOf} properties.

@@

The commonality of all quality checking functions is that they take a given SKOS vocabulary as input and return a (nested) set, containing the resources fulfill the quality criterion. Therefore, we can formalize a quality criterion as being an abstract quality checking function $f:V \rightarrow \mathcal{P}(IR)$, mapping a given vocabulary $v \in V$ to the powerset of its resources. Each quality criterion we found can be implemented as a realization of such an abstract filtering function, possibly with additional parameters. Statistical analysis can be computed on top of the return values.

@@

SKOS defines a set of properties that link resources with RDF Literals, which are plain text natural language strings with an optional language tag. This includes the labeling properties \texttt{skos:prefLabel}, \texttt{skos:altLabel}, \texttt{skos:hiddenLabel} and also SKOS documentation properties, such as \texttt{skos:note} and subproperties thereof. Since omitting language tags or using non-standardized, private language tags in a SKOS vocabulary could limit language-dependent structural queries, we check a SKOS vocabulary for omitted language tags by iterating over resources in $IR$ and collect those that have relations to plain literals without or with invalid language tags. We delegate validity checking of language tags to methods provided by the by the OpenRDF\footnote{\url{http://www.openrdf.org/}} framework that in turn checks for conformance with RFC3066\footnote{\url{http://www.ietf.org/rfc/rfc3066.txt}}.\todo{CM}{is there any literature reference for this criterion?}

@@

A SKOS vocabulary that uses language tags with its literal values, should apply them consistently for all concepts. Hence, the set of language tags used by the literal values linked with a concept should be the same for all concepts. This is particularly important for applications that rely on internationalization and translation use cases. We calculate the affected concepts by extracting the global set of language tags used in a vocabulary from all literal values attached to all concepts in $C$. In a second pass we again iterate over this set of concepts and identify all conceptual resources having a set of language tags that is not equal to the global language set. \todo{CM}{literature reference?}

@@

\cite{Svenonius1997} advocates the ``inclusion of as much definition material as possible''. The SKOS language specification defines various properties that are intended to hold this kind of information, subsumed as ``documentation properties''~\cite{SkosReference2008}. To identify all undocumented concepts, we iterate over all concepts in $C$ and collect those for which we cannot find information linked by any of the documentation properties.

@@

This criterion is an extension of a statement from \cite{Isaac2009}, recommending ``no two concepts have the same preferred lexical label in a given language when they belong to the same concept scheme''. Based on our review of existing vocabularies, we could observe that (i) many vocabularies don’t make use of concept schemes and (ii) also identical \texttt{skos:altLabels} or \texttt{skos:hiddenLabels} can have a negative effect in some use-cases (e.g, auto-completion of concept labels based on user input). Thus we follow the more general approach of finding all pairs of concepts with their respective \texttt{skos:prefLabel}, \texttt{skos:altLabel} or \texttt{skos:hiddenLabel} meeting a threshold of a certain similarity function $sim:ULV \times ULV \rightarrow [0,1]$. For the analysis of the criterion in this document, we define the similarity function to check for case-insensitive string equality and the threshold to equal 1. Calculation of this criterion is done by iterating over all pairs of concepts, i.e., elements of $C \times C$. We then apply the $sim$ to every possible combination of labels between the members of the pairs and collect those pairs for which $sim$ meets or exceeds a specified threshold.

>>>>

\begin{definition}[SKOS Vocabulary] Let a SKOS vocabulary be a tuple of the form $V = \langle IR, C, SR, ULV \rangle$, with \(IR = I_{CEXT}(\texttt{rdfs:Resource}^\mathcal{I})\) being the set of \textbf{resources}, \(C = I_{CEXT}(\texttt{skos:Concept}^\mathcal{I})\) being the set of \textbf{conceptual resources}, \(SR = I_{EXT}(\texttt{skos:semanticRelation}^\mathcal{I})\) being the set of \textbf{semantic relations} associating conceptual resources with one another, and $ULV \subseteq I_{CEXT}(\texttt{rdfs:Literal}^\mathcal{I})$ being the set of \textbf{untyped literal values}. Further, we let $V$ be the fully entailed RDFS interpretation~\cite{RDFSEM2012} of the underlying RDF graph, enriched by entailment of \texttt{owl:inverseOf} properties.

@@

% @CM Are you sure about the term "untyped literal"? The RDF spec talkes about
% plain literals

The commonality of all quality checking functions is that they take a given SKOS vocabulary as input and return a (nested) set, containing the resources \todo{CM}{fulfill the quality criterion?}. Therefore, we can formalize a quality criterion as being an abstract quality checking function $f:V \rightarrow \mathcal{P}(IR)$, mapping a given vocabulary $v \in V$ to the powerset of its resources. Each quality criterion we found can be implemented as a realization of such an abstract filtering function, possibly with additional parameters. Statistical analysis can be computed on top of the return values.\todo{CM}{the term criterion still sounds strange. We must somehow clarify what passing/failing a criterion means there.}

@@

SKOS defines a set of properties that link resources with RDF Literals, which are plain text natural language strings with an optional language tag. This includes the labeling properties \texttt{skos:prefLabel}, \texttt{skos:altLabel}, \texttt{skos:hiddenLabel} and also SKOS documentation properties, such as \texttt{skos:note} and subproperties thereof. Since omitting language tags or using non-standardized, private language tags in a SKOS vocabulary could limit language-dependent structural queries, we check a SKOS vocabulary for omitted language tags. We iterate over all resources in $IR$ and collect those that have relations to plain literals with missing or invalid language tags, hence tags that are not defined in RFC3066\footnote{RFC 3066: Tags for the Identification of Languages \url{http://www.ietf.org/rfc/rfc3066.txt}}.

% This is implementation specific; let's stay with high-level descriptions of the algorithms
%We delegate validity checking of language tags to methods provided by the by the OpenRDF\footnote{\url{http://www.openrdf.org/}} framework that in turn checks for conformance with RFC3066\footnote{\url{http://www.ietf.org/rfc/rfc3066.txt}}.\todo{CM}{is there any literature reference for this criterion?}

@@

A SKOS vocabulary that uses language tags with its literal values should apply them consistently for all concepts. Hence, the set of language tags used by the literal values linked with a concept should be the same for all concepts. This is particularly important for applications that rely on internationalization and translation use cases. We calculate the affected concepts by extracting the global set of language tags used in a vocabulary from all literal values attached to all concepts in $C$. In a second pass we again iterate over this set of concepts and identify all conceptual resources having a set of language tags that is not equal to the global language set. \todo{CM}{literature reference?}

@@

\cite{Svenonius1997} advocates the ``inclusion of as much definition material as possible''. The SKOS language specification defines various properties that are intended to hold this kind of information, subsumed as ``documentation properties''~\cite{SkosReference2008}. To identify all undocumented concepts, we iterate over all concepts in $C$ and collect those that do not use any of these documentation properties.

@@

This criterion is an extension of a statement from \cite{Isaac2009} that recommends that ``no two concepts have the same preferred lexical label in a given language when they belong to the same concept scheme''. Based on our review of existing vocabularies, we could observe that (i) many vocabularies don’t use concept schemes and (ii) identical \texttt{skos:altLabels} or \texttt{skos:hiddenLabels} can also negatively affect use cases such as auto-completion, which proposes labels based on the user input. Thus, this criterion generalizes this statement by finding all concept pairs with their respective \texttt{skos:prefLabel}, \texttt{skos:altLabel} or \texttt{skos:hiddenLabel} property values meeting a certain similarity threshold defined by a function $sim:ULV \times ULV \rightarrow [0,1]$. The default, built-in similarity function checks for case-insensitive string equality with a threshold equal to 1. This criterion is calculated by iterating over all pairs of concepts, i.e., elements of $C \times C$, applying $sim$ to every possible label combination, and collecting those pairs for which $sim$ meets or exceeds a specified threshold.
